---
title: "Covid91 vaccine study FinalSpring2023"
author: "Andrew Valles Tovar"
output: 
  html_document:
    number_sections: true
    toc: true
    toc_depth: 5
date: "`r Sys.Date()`" 
---

```{r include=FALSE,echo=FALSE}
require(tidyverse)
require(tigerstats)
require(forcats)
FinalDat <-read.csv(file="finalFall2022.csv",header=TRUE)
#recode to remove the annoying "X" from the dataframe
FinalData <- FinalDat  %>% select(-one_of("X"))
FinalData <- FinalData %>%
mutate(across(where(is.character), as.factor))
male = filter(FinalData, sex =="M")
female = filter(FinalData, sex =="F")
LGBTQ = filter(FinalData, LGBTQ =="gay")
druggie = filter(FinalData, DrugUser =="yes")
```

# Introduction

A new  preventative vaccine (covid921) is under test in clinical trials. Given the nature of this vaccine it is suspected that the drug may exhibit different effectiveness across the variety of subpopulations tracked. Consider population subgroups by gender( guys and gals), drug use, and sexual orientation. Write a complete report that separately considers this drug for each subgroup and your recommendations as to whether people in the subgroup should take the vaccine. Structure your report in four sections each devoted to one of the four subgroups, use null and alternative hypothesis statements and correct statistical verbage when referencing p-values in conclusions.

# Coding details

To tackle this study, we need to separate males, females, gay people, and drug users into four subsections using our clinical trial data named "FinalData". Doing this will allow us to go in-depth in whether or not the vaccine is an effective treatment on the specific subsections.

To do this we use the following R code to separate the aforementioned population into a data set.

name = filter(FinalData, category =="x")

What this R code does is in our clinical trial data named "FinalData", we are filtering it for a specific category and value and creating a data set for it. If we are looking for female participants, we would replace "category" with "sex" and have it equal to "F" which represents females.

Now with our four newly created data sets, we can start taking a look to see if the vaccine is effective for each of the data sets. 

## Males

### Bar Chart
```{r}
barchartGC(~infected + treatment,data=male)
barchartGC(~infected + treatment,data=male, type="percent")
```

There are two bar plots here. The one we'll talk about first is the percent bar plot which shows the percent of males in the population that took either the placebo or drug and whether or not they caught COVID. Looking at the percent bar plot, it seems a substantial amount of infected males were under the effects of the placebo. It seems about 65% of infected males took the placebo, and around 35% of took the vaccine. So far it seems the vaccine is effective in preventing the disease for males.

The second plot is almost the same as the one we explained, but instead the y-axis represents a count of the male population instead of percentage. As we can see, only a small amount of males in the population actually contracted COVID-19 compared to the section that didn't catch it. The bar plots going forward from here follow the same explanation.

### Numerical Descriptive Results
```{r}
table1 <- xtabs(~infected + treatment, data=male)
rowPerc(table1)
colPerc(table1)
```
The numerical results here show the exact percentage of males taking either the placebo or vaccine that caught COVID-19 and the percentage of males in the population that contracted the disease. The numerical results going forward from here follow the same explanation.

The row table shows of those infected with COVID-19, 68% of males had taken the placebo and only 32% took the vaccine. Once again, it seems the vaccine is proving to be effective in preventing the disease for males. It also seems like there were more males that took the vaccine rather than the placebo.

The column table shows the percentage of males in the population that either took the placebo or vaccine and whether or not they caught the disease in their respective category. So nearly 3% of males that took the placebo contracted the disease and a mere 1% of males that took the vaccine were infected.

### Inferential Results
```{r}
chisq.test(table1)
chisqtestGC(table1)
malesfisher <- fisher.test(table1)
malesfisher
```
The inferential results here show the exact amount of males in the population and whether or not they caught COVID-19 with the placebo or vaccine.

245 males that had taken the placebo got COVID-19 and 113 of males that took the vaccine also caught the disease. Since there is a difference of 132 males, it seems the vaccine is effective in preventing the disease.

The null hypothesis here is there is no relationship between the treatment and males and the alternative hypothesis is there is a relationship between the treatment and males. The P-value indicated by Fisher's test is 0.000000000002364 which is lower than 0.05. The P-value indicated by the Chi-square test is 0.000000000005307 which is also below 0.05. Therefore, we reject the null in favor of the alternative.

The odds ratio indicated by Fisher's test is 2.2.

```{r}
effect <- malesfisher$p.value/(malesfisher$p.value +1)
effectivenessmale = malesfisher$estimate/(malesfisher$estimate +1)*100
print(effectivenessmale)
```

```{r}
low=(malesfisher$conf.int[1]/(malesfisher$conf.int[1] +1))*100
hi=(malesfisher$conf.int[2]/(malesfisher$conf.int[2] +1))*100
options(digits=2)
```

### Effectiveness Confidence Interval `r low`  to `r hi`

Shown above is the effectiveness confidence interval for males. If the interval isn't above 50%, then that means the vaccine is not effective in protecting males against COVID-19. The confidence level is shown as `r low`  to `r hi` meaning the vaccine is a success for males. To measure how effective the vaccine is on males, we use the odds ratio. The odds ratio here is listed as `r effectivenessmale` meaning it is effective.

## Females

### Bar Chart
```{r}
barchartGC(~infected + treatment,data=female)
barchartGC(~infected + treatment,data=female, type="percent")
```

Strangely, it seems like about 58% of females that contracted COVID-19 had taken the vaccine and about 42% took the placebo. Whereas the vaccine seemed effective on males, this doesn't seem like the case for females. On the frequency graph, only a small amount of females in the population contracted the disease.

### Numerical Descriptive Results
```{r}
table1 <- xtabs(~infected + treatment, data=female)
rowPerc(table1)
colPerc(table1)
```
Of the females infected with COVID-19, the row table here shows 57% took the vaccine and around 43% took the placebo. There also seems to be slightly more placebo participants than the vaccine. Once again it doesn't seem like the vaccine is helping females very much.

For the column table, 1% of females that took the placebo caught the disease and 2% took the vaccine. So while it is a rather small percentage, more females contracted COVID-19 with the vaccine.

### Inferential Results
```{r}
chisq.test(table1)
chisqtestGC(table1)
fisher.test(table1)
```
A total of 207 females that contracted COVID-19 took the vaccine and 155 took the placebo. There's a difference of 57 females, so it doesn't seem like the vaccine is very effective on them. 

The null hypothesis here is there is no relationship between the treatment and females and the alternative hypothesis is there is a relationship between the treatment and females. The P-value indicated by Fisher's test is 00.004887 and the P-value indicated by the Chi-square test is 0.005747. Both of the P-values are below 0.05. Therefore, we reject the null in favor of the alternative.


## LGBTQ

### Bar Chart
```{r}
barchartGC(~infected + treatment,data=LGBTQ)
barchartGC(~infected + treatment,data=LGBTQ, type="percent")
```

The percent bar plot here shows some interesting results. From those that were infected and answered yes to being gay, about 82% took the vaccine and about 18% took the placebo. It seems like if you're LGBTQ, you're highly likely to contract COVID-19 than of the placebo.

For the frequency plot, only a tiny amount of individuals caught the disease whether or not they took the vaccine or placebo with the vaccine being the leading cause.

### Numerical Descriptive Results
```{r}
table1 <- xtabs(~infected + treatment, data=LGBTQ)
rowPerc(table1)
colPerc(table1)
```
Starting with the row table, we can see a surprising amount of LGBTQ's infected with COVID-19 had taken the vaccine. 87% of the ones infected were distributed the vaccine and 13% were given the placebo. 87% is quite the number so it'd be hard to argue that the vaccine was a success for them.

For the column table, it shows 3% of LGBTQ's given the vaccine caught COVID-19 and not even 1% that were given the placebo contracted the disease.

### Inferential Results
```{r}
chisq.test(table1)
chisqtestGC(table1)
fisher.test(table1)
```
According to the inferential data, 52 LGBTQ's infected with COVID-19 took the vaccine and only 8 were given the placebo. Unfortunately, it doesn't seem like the vaccine does it's job if you're LGBTQ.

The null hypothesis here is there is no relationship between the treatment and LGBTQ's and the alternative hypothesis is there is a relationship between the treatment and LGBTQ's. The P-value indicated by Fisher's test is 0.00000001244 and the P-value indicated by the Chi-square test is 0.0000001104. Since both of the P-values are below 0.05, we reject the null in favor of the alternative.

## Druggies

### Bar Chart
```{r}
barchartGC(~infected + treatment,data=druggie)
barchartGC(~infected + treatment,data=druggie, type="percent")
```

The bar plot above represents the population of participants that answered to yes drug use and if they were infected while under the effects of the placebo or vaccine. We can see that in this case as well, the vaccine doesn't seem to be working if you're a drug user. About 86% of those infected had taken the vaccine and about 14% took the placebo.

For the frequency bar plot, once again it seems that only a small amount of drug user participants actually contracted COVID-19 compared to the OK section.

### Numerical Descriptive Results
```{r}
table1 <- xtabs(~infected + treatment, data=druggie)
rowPerc(table1)
colPerc(table1)
```

Looking at the row table, we can see that 83% of drug users that contracted COVID-19 had taken the vaccine and almost 17% took the placebo. We can also see that there were slightly more vaccine participants than placebo participants for drug users.

The column table here shows 3% of drug users that were given the vaccine got infected with COVID-19 and barely 1% had taken the placebo. So if you're a drug user, it is more likely to contract the disease while taking the vaccine.

### Inferential Results
```{r}
chisq.test(table1)
chisqtestGC(table1)
fisher.test(table1)
```

With the inferential results displayed here, a total of 50 drug users undertaking the vaccine were infected with COVID-19 and 10 were given the placebo. So it's more likely to contract COVID-19 with the vaccine than with the placebo.

The null hypothesis here is there is no relationship between the treatment and drug users and the alternative hypothesis is there is a relationship between the treatment and drug users. The P-value indicated by Fisher's test is 0.0000003119 and the P-value indicated by the Chi-square test is 0.000001419. Both of these P-values are below 0.05, so we reject the null in favor of the alternative.


# Overall Results and Conclusions